A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-21 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reviewed capacities

\textbf{Probability of outcomes vs fraction of components.} From the previous discussion, we can see that the problem with using probability in quantum mechanics is that not all pure states represent mutually exclusive events, and it is only over mutually exclusive events that classical probability can be defined. Mutually exclusive events correspond to ensemble are orthogonal, as each instance of one ensemble cannot be produced by the other ensemble. That is why, in quantum mechanics, probability can be defined only once a context, an orthogonal basis, is chosen. Therefore, at a conceptual level, probabilities can only give us a patchwork representation of ensembles. Some author embrace this view, by characterizing states as the collection of all probability distributions of all possible results, as it is generally the case in GPTs. [CITATION NEEDED] However, there is a different route.

Note that the linearity in our axioms is in terms of mixing coefficients, not probability of outcomes. That is, regardless of whether we can characterize experiment outcomes with probability distributions, preparation procedures will always allow us to create statistical mixture. We can create statistical mixtures regardless of whether the states we are mixing represent mutually exclusive outcomes or not. That is, instead of focusing on the likelihood of an outcome, we will focus on how much of a particular ensemble can be understood as the mixture of some other ensembles. Instead of focusing on what an ensemble may cause, we focus on how an ensemble can be produced. It turns out, this second approach generalizes in a natural way.

@@

Note that, if not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure. It is therefore a non-additive probability measure. The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature.\footnote{ \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} } We will see how our approach fits  in that literature in a later section.

\textbf{Compare to Choquet theory.} There is a straight forward link between our non-additive measure and Choquet theory. Recall that Choquet theory start by defining measure on the whole ensemble space before finding those constrained to the extreme points. Our setup is similar. The fraction capacity is defined on the whole ensemble space. In the classical discrete case, for example, if $A \subset \Omega \subset \Ens$ is an event, then $\frcap_{\ens}(A)$ will return the probability of event $A$; if $P \subset \Ens$ represents all the Poisson distributions with integer coefficient, $\frcap_{\ens}(P)$ will return how much of $\ens$ can be expressed as a mixture of Poisson distributions. In case that extreme points exist, or we can extend the ensemble space with some limit procedure, we can also define the fraction capacity on the extreme points.

We can establish a direct connection between the fraction capacity and Choquet theory, because the ability to express an ensemble $\ens$ as the convex combination of $A$ is precisely the ability to find a measure $\mu$ that has support within $A$. That is, $\ens = \sum_{i=1}^{\infty} p_i \ens[a]_i$ if an only if the probability measure defined by $\mu(\{\ens[a]_i\}) = p_i$ represents $\ens$.\footnote{TODO: add proof somewhere} Consequently, if $M_{\ens}$ is the set of all probability measures that represent $\ens$ as defined in Choquet theory, then $\frcap_{\ens}(A) = \sup(\{\mu(A) \, | \, \mu \in M_{\ens}\})$.\footnote{TODO: prove somewhere} The fraction capacity is simply the supremum on each set over all possible measures.

As we saw, the classical case corresponds to a simplex in Choquet theory. This means that each element of the ensemble space is represented by a unique probability measure over the extreme points. In that case, the fraction capacity restricted to the extreme points will also be additive. In the non-classical case, the closure of the ensemble space will not be a simplex, and there will be multiple representations, which will lead to a non-additive measure. Suppose, in fact, that $\mu$ and $\nu$ are two distinct measures that represent the same ensemble $\ens$. Then there will be a set $A$ such such that $\mu(A) \neq \nu(A)$. Suppose, without loss of generality, that $\mu(A) > \nu(A)$ We will also have $\mu(A^{\complement}) = 1- \mu(A) < 1 - \nu(A) = \nu(A^{\complement})$. Since the fraction capacity for a set cannot be less than the measure on that set, we have $\frcap_{\ens}(A \cup A^{\complement})=\frcap_{\ens}(X) = 1 = \mu(A) + \mu(A^{\complement}) < \mu(A) + \nu(A^{\complement}) \leq \frcap_{\ens}(A) + \frcap_{\ens}(A^{\complement})$. That is, the fraction capacity is additive over disjoint sets of extreme points in and only in the classical case. In other words, the fraction capacity reduces to a classical probability distribution over pure states in the classical case.

\textbf{Uniqueness of the fraction capacity.} From a conceptual standpoint, the fraction capacity has a major advantage: it has a clear physical meaning and it is unique. Like in Choquet theory, and unlike for pseudo-probability, the domain over which the measure is defined is fully determined by the system, by the ensemble space. However, unlike Choquet, and like pseudo-probability once the space is fixed, there one unique fraction capacity for each ensemble. Moreover, the setup is more general than Choquet theory (i.e. the ensemble space need not be compact) and stems from necessary physical requirement.  An in depth study on fraction capacities, then, is guaranteed to be applicable to all physical theories, including those yet to be formulated.

@@

To build an intuition, consider the classical case. For every a region $U$ of phase space, we have a set of possible probability distributions that have support $U$. This is a convex set since every mixture of distributions with support $U$ has support $U$. Note that the uniform distribution $\rho_U$, if it exists, is the one with the greatest entropy, which will correspond to $S(\rho_U) = \log \mu(U)$. If $U$ is not compact, then there is no upper limit to the entropy and $\mu(U) = +\infty$. Now, let us transfer this intuition on the ensemble space $\ens$. If we take a set $A$, we can ask: what is the highest entropy reachable by a mixture of elements of $A$? If $A$ happens to be the set of ensembles with support $U$, then the supremum of the entropy is $\log \mu(U)$, which works also in the unbounded case. But the setup now is more general: $A$ can be any set of ensembles in any ensemble space.

We still need to make sure that this definition still makes sense physically. If the entropic volume measure the spread of an ensemble, during a mixture the spread can at most sum. Remarkably, the bounds on the entropy are such that the exponential of the entropy is subadditive. That is, $2^{S(p\ens[a]+\bar{p}\ens[b])} \leq 2^{S(\ens[a])} + 2^{S(\ens[b])}$ for any ensemble space.\footnote{It is an open question whether the exponential is the only monotonic function of the entropy that is subadditive.} Therefore this is indeed a general way to define the spread of ensembles and therefore the count of distinguishable states.

@@

Again we note that the non-additive measure we recover is very well-behaved, and connects directly with important quantities. When Choquet theory applies, state capacity can be extended to the extreme points. In the classical case, disjoint sets of extreme points will be orthogonal, meaning that the state capacity there restricted is an additive measure. In the discrete case, every point is associated to zero entropy, and since $2^0=1$, the state capacity reduces to the counting measure. In the continuous case, gives a subset of phase space $U$ and $A_U$ the set of distributions with support within $U$, the maximum entropy is reached by the uniform distribution $\rho_U \in A_U$, which means we recover $\log_2 \capacity(A_U) = S(\rho_U)$, the fundamental postulate of statistical mechanics. Similarly, in quantum mechanics, given an orthogonal subspace $A$, the maximum entropy is recovered by the maximally mixed state, which means $\capacity(A) = 2^{\dim(A)}$.

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable, the measure is additive. In quantum mechanics, only orthogonal state are distinguishable, therefore the measure is non-additive over pure states. We see again how the lack of single decomposition of ensembles over pure states is responsible for the failure of the additivity of the measure.

>>>>

\textbf{Probability of outcomes vs fraction of components.} We saw that classical probability fails in quantum mechanics because pure states are not always mutually exclusive, are not orthogonal. That is why, in quantum mechanics, probability can be defined only once a context, an orthogonal basis, is chosen. Therefore, at a conceptual level, probabilities can only give us a patchwork representation of ensembles. Some author embrace this view, by characterizing states as the collection of all probability distributions of all possible results, each distribution providing partial information. [CITATION NEEDED] However, there is a different route.

Note that the linearity in our axioms is in terms of mixing coefficients, not probability of outcomes. That is, regardless of whether we can characterize experiment outcomes with probability distributions, preparation procedures will always allow us to create statistical mixture. As we saw in Choquet theory, the difficulty is that there are multiple was to mix pure states to recover the same ensemble. However, the question of what is the biggest mixture coefficient one can use in a specific case is well-defined, and generalizes to the non-compact case as well. Therefore, our generalization of probability is not in terms of what an ensemble can lead to (i.e. measurements), but in terms of how ensembles can be produced (i.e. preparations).

@@

Note that, if not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure. It is therefore a non-additive probability measure. The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature.\footnote{ \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} }

\textbf{Compare to Choquet theory.} In case $E$ is compact, we can restrict the fraction capacity to the set $X$ of extreme points. Given a set $A \subset X$, of extreme points, we can find the set $N \subset E$ of all ensemble that can be represented with a probability measure with support $A$. This forms a convex set. Now, the ability to express an ensemble $\ens$ as the convex combination of $N$ is precisely the ability to find a measure $p$ that represents $\ens$ such that $p(A) \neq 0$. That is, $\ens = \lambda \ens[a] + \bar{\lambda} \ens[b]$ with $\ens[a] \in M$ if and only if there is a probability measure $p$ that represents $\ens$ such that $p(A) \geq \lambda$.\footnote{TODO: add proof somewhere} Consequently, if $M_{\ens}$ is the set of all probability measures that represent $\ens$ as defined in Choquet theory, then $\frcap_{\ens}(A) = \sup(\{p(A) \, | \, p \in M_{\ens}\})$.\footnote{TODO: prove somewhere} The fraction capacity is simply the supremum on each set over all possible measures.

As we saw, the classical case corresponds to a simplex in Choquet theory. This means that each element of the ensemble space is represented by a unique probability measure over the extreme points. In that case, the fraction capacity restricted to the extreme points will also be additive. In the non-classical case, the closure of the ensemble space will not be a simplex, and there will be multiple representations, which will lead to a non-additive measure. Suppose, in fact, that $p_1$ and $p_2$ are two distinct measures that represent the same ensemble $\ens$. Then there will be a set $A$ such such that $p_1(A) \neq p_2(A)$. Suppose, without loss of generality, that $p_1(A) > p_2(A)$ We will also have $p_1(A^{\complement}) = 1- p_1(A) < 1 - p_2(A) = p_2(A^{\complement})$. Since the fraction capacity for a set cannot be less than the measure on that set, we have $\frcap_{\ens}(A \cup A^{\complement})=\frcap_{\ens}(X) = 1 = p_1(A) + p_1(A^{\complement}) < (A) + p_2(A^{\complement}) \leq \frcap_{\ens}(A) + \frcap_{\ens}(A^{\complement})$. That is, the fraction capacity is additive over disjoint sets of extreme points in and only in the classical case. In other words, the fraction capacity reduces to a classical probability distribution over pure states in the classical case.

\textbf{Uniqueness of the fraction capacity.} From a conceptual standpoint, the fraction capacity has a major advantage: it has a clear physical meaning and it is unique. Like in Choquet theory and unlike for quasi-probability, the domain over which the measure is defined is fully determined by the system, by the ensemble space. On the other hand, unlike Choquet theory and like quasi-probability, once the space is fixed, there is one unique fraction capacity for each ensemble. Moreover, the setup is more general than Choquet theory (i.e. the ensemble space need not be compact) and stems from necessary physical requirement. An in depth study on fraction capacities, then, is guaranteed to be applicable to all physical theories, including those yet to be formulated.

@@

To build an intuition, consider the classical case. For every region $U$ of phase space, we have a set of possible probability distributions that have support $U$. This is a convex set since every mixture of distributions with support $U$ has support $U$. Note that the uniform distribution $\rho_U$, if it exists, is the one with the greatest entropy, which will correspond to $S(\rho_U) = \log \mu(U)$. If $U$ is not compact, then there is no upper limit to the entropy and $\mu(U) = +\infty$. Now, let us transfer this intuition on the ensemble space $\ens$. If we take a set $A$, we can ask: what is the highest entropy reachable by a mixture of elements of $A$? If $A$ happens to be the set of ensembles with support $U$, then the supremum of the entropy is $\log \mu(U)$, which works also in the unbounded case.

The above intuition started with the notion of the state space $X$ and the count of states $\mu$. In the general case, we will start with the ensemble space $\Ens$ and the entropy $S$. For ensembles, we can define the entropic volume\footnote{CITE Hall} to be the exponential of the entropy $2^{S(\ens)}$. For the entropic volume to quantify the spread of the ensemble, we must have that the spread can at most sum. Remarkably, the bounds on the entropy are such that the exponential of the entropy is subadditive. That is, $2^{S(p\ens[a]+\bar{p}\ens[b])} \leq 2^{S(\ens[a])} + 2^{S(\ens[b])}$ for any ensemble space.\footnote{It is an open question whether the exponential is the only monotonic function of the entropy that is subadditive.} For the equality to hold, the ensemble must be orthogonal, meaning that two states are distinguishable if and only if they are mutually exclusive. Therefore, this is indeed a general way to define the spread of ensembles and the count of distinguishable states in a way that is consistent with the previous discussions.

@@

Again we note that the non-additive measure we recover is very well-behaved, and connects directly with important quantities. In the compact case, state capacity can be extended to the extreme points. In the classical case, disjoint sets of extreme points will be orthogonal, meaning that the state capacity there restricted is an additive measure. In the discrete case, every point is associated to zero entropy, and since $2^0=1$, the state capacity reduces to the counting measure. In the continuous case, gives a subset of phase space $U$ and $A_U$ the set of distributions with support within $U$, the maximum entropy is reached by the uniform distribution $\rho_U \in A_U$, which means we recover $\log_2 \capacity(A_U) = S(\rho_U)$, the fundamental postulate of statistical mechanics. Similarly, in quantum mechanics, given an orthogonal subspace $A$, the maximum entropy is recovered by the maximally mixed state, which means $\capacity(A) = 2^{\dim(A)}$.

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable, the measure is additive. In quantum mechanics, only orthogonal state are distinguishable, therefore the measure is non-additive over pure states. We see again how the lack of single decomposition of ensembles over pure states, which equates to the lack of mutually exclusiveness, is responsible for the failure of the additivity of the measure.
